---
title: "DB-Biostatistics Draft Major Project"
author: "Derek Bornath"
date: "16/03/2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlme)
library(car)
library(ez)
library(afex)
library(ggplot2)
library(emmeans)
library(gvlma)
```

#Starting by reading in the data set
```{r}
EMRL=read.csv('EMRL.Ghr.GLP-1.App.csv')
View(EMRL)
```


##Running an anova to determine if there are any differences between menstrual phase and appetite surpressing hormone GLP-1
```{r}
EMRL.anova2=aov_ez(data=EMRL,
  id= "Participant",
  within = c("Phase","Time"),
  dv="GLP.1",
  type="3"
)
summary(EMRL.anova2)
```
#After we will run a post hoc test to determine where the interactions occur.
```{r}
con.GLP.1 <- lsmeans(EMRL.anova2$aov,~Phase|Time)
con.GLP.1 <- lsmeans(EMRL.anova2$aov,~Time*Phase)

con.GLP.1
contrast(con.GLP.1,method="pairwise")

```

##Lastly we will determine if there is differences between menstrual phase and overall appetite perception
```{r}
EMRL.anova3=aov_ez(data=EMRL,
  id= "Participant",
  within = c("Phase","Time"),
  dv="Appetite",
  type="3"
)
summary(EMRL.anova3)
```

#This will be followed again by a post hoc test to determine where interactions occur.
```{r}
con.App <- lsmeans(EMRL.anova3$aov,~Phase|Time)
con.App <- lsmeans(EMRL.anova3$aov,~Time*Phase)

con.App
contrast(con.App,method="pairwise")

```

#Next we will create an interaction plot to depict these analyses.
```{r}
interaction.plot(EMRL$Time,EMRL$Phase,EMRL$GLP.1, xlab = 'Exercise Time Points', ylab = 'GLP-1 Concentration (pM/mL-1)')
interaction.plot(EMRL$Time,EMRL$Phase,EMRL$Appetite, xlab = 'Exercise Time Points', ylab = 'Overall Appetite Perception (mm)')
```

##Since the interaction plots are not easy to read or make connections, we have also completed ggplot histograms to better depict the differences
```{r}
ggplot(data=EMRL, aes(x=as.factor(Time), y=GLP.1, fill=Phase)) + 
  geom_histogram()+
  xlab("Exercise Time Points")+
  ylab("Ghrelin (pg/mL-1)")+
  theme(panel.background = element_rect(fill = "lightcyan3", colour = "lightcyan3",size = 0.5, linetype = "solid"),panel.grid.major = element_line(size = 0.5, linetype ='solid', colour = "white"), panel.grid.minor = element_line(size = 0.25, linetype ='solid',colour = "white"), axis.text=element_text(size=12), axis.title=element_text(size=10,face="bold"),legend.title=element_text(size=12), legend.text=element_text(size=10))+
  theme(legend.position = c(0.2,0.9))
```


##Lastly we will test the assumptions.#Test assumptions of rm-ANOVA

#Assumptions for a repeated-measures ANOVA:
#1. Independence of samples 
#2. Homogeneity of variances among treatments
#3. Normality of residuals
#4. Sphericity

#The first assumption can only be validated by understanding how samples were collected, and therefore depends on good experimental design. 

#The second assumption testing for homogeneity of variances, we can use Levene's test
```{r}

leveneTest(y=EMRL$GLP.1,group=EMRL$Time) # by Time

leveneTest(y=EMRL$GLP.1,group=EMRL$Phase) # by Phase

leveneTest(y=EMRL$Appetite,group=EMRL$Time) # by Time

leveneTest(y=EMRL$Appetite,group=EMRL$Phase) # by Phase
```

#The third assumption testing for normality of residuals by performing a Shapiro-Wilks test and looking at the distribution through a histogram.
```{r}
shapiro.test(residuals(EMRL.anova2$lm)) # for GLP-1
hist(residuals(EMRL.anova2$lm))

shapiro.test(residuals(EMRL.anova3$lm)) # for Appetite
hist(residuals(EMRL.anova3$lm))
```

#Assumption 4 was tested through each ANOVA output.


##The other statistical Analysis of my data will be a linear regression to determine any predictor variables in the data. I am reading in the same data again, just organized differently to allow for linear regressions.
```{r}
EMRL.Wide=read.csv('EMRL.Ghr.GLP-1.App.Wide.csv')
View(EMRL.Wide)
```

#Scatter Plots to compare appetite to appetite hormones immediately Post-Exercise
```{r}
plot(EMRL.Wide$FP.APP.Imm.Post ~EMRL.Wide$FP.GLP.1.Imm.Post, main='Comparison of Follicular Phase Appetite to GLP-1 Concentration Post-Exercise', ylab='GLP-1 Concentration (pM/mL-1)', xlab= 'Overall Appetite (mm)')

plot(EMRL.Wide$LP.APP.Imm.Post ~EMRL.Wide$LP.GLP.1.Imm.Post, main='Comparison of Luteal Phase Appetite to GLP-1 Concentration Post-Exercise', ylab='GLP-1 Concentration (pM/mL-1)', xlab= 'Overall Appetite (mm)')
```

#Performing a linear regression to determine if appetite hormones have an impact on appetite perceptions
```{r}
FP.App.GLP.1.comp1=glm(EMRL.Wide$FP.APP.Imm.Post ~EMRL.Wide$FP.GLP.1.Imm.Post)
summary.lm(FP.App.GLP.1.comp1)

LP.App.GLP.1.comp1=glm(EMRL.Wide$LP.APP.Imm.Post ~EMRL.Wide$LP.GLP.1.Imm.Post)
summary.lm(LP.App.GLP.1.comp1)
```

#adding the regression line determined to the scatter plots previously made
```{r}
plot(EMRL.Wide$FP.APP.Imm.Post ~EMRL.Wide$FP.GLP.1.Imm.Post, ylab='GLP-1 Concentration (pM/mL-1)', xlab= 'Overall Appetite (mm)')
abline(FP.App.GLP.1.comp1, col=c('blue'))

plot(EMRL.Wide$LP.APP.Imm.Post ~EMRL.Wide$LP.GLP.1.Imm.Post, ylab='GLP-1 Concentration (pM/mL-1)', xlab= 'Overall Appetite (mm)')
abline(LP.App.GLP.1.comp1, col=c('blue'))
```

#Using the gvlma function to determine if the assumptions of a linear regression were met.
```{r}
gvlma(lm(FP.App.GLP.1.comp1))

gvlma(lm(LP.App.GLP.1.comp1))

```
